Stage 1 Registered Report: The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-05 article.tex
Exploratory

This exploratory research's goal is to produce a data set that 
quantifies the inference error we make from choosing an overly
simplistic prior. In this context, we measure the error made on
phylogenies in which speciation takes time, when using a Birth-Death
prior.

To quantify the inference error, we simulate 'true' incipient species
trees, sample a 'true' species trees from those, simulate a DNA alignments 
that follows a species tree's ontogeny, then infer a posterior of
phylogenies from each alignment, using an overly simplistic Birth-Death
prior. The 'true' species tree is compared to
the posterior's phylogenies with the nLTT statistic, which can be viewed
as a measure of similarity. We predict this error to increase with an
increasing protractednedness (that is, a low speciation completion rate),
but we cannot predict the the extent of this error, as it has never been 
measured.

The data set produced by the exploratory reseach will consist out of two files:
one containing raw data, the other summarized data. 
A data file should be convenient to handle, thus as an upper limit for this data set's size, 
we use the maximum size of a table (called a data frame) of the R programming 
language. Within R, such a table may have maximally $2^{32}-1$ cells and occupy
less then 2 gigabytes in memory. 
The raw data file will contain parameters and 1000 nLTT statistics values.
Each row will contain the parameter set and these nLTT statistics.
From the constraints of the R programming language, this allows for 40K rows. 
The file with the summarized data will provide summary statistics of the raw
data. Each row will contain a parameter set, summary statistics and other
measurements. See the appendix, chapter 'Data set specifications', for the full list.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We simulate protracted birth-death trees. In a protracted birth-death tree,
species-to-be are present, but not yet recognized as different from
their ancestral species. We simulate these trees from the biological parameters 
in a factorial fashion. Excluded from this are combinations of parameters in which
speciation initiation rate is lower or equal the extinction rate, as these
are biologically irrelevant. Out of computational necessity, 
parameter combinations that result in an expected number of taxa above 1000 
are removed as well.

From these PBD trees, we sample species tree in two different ways. 
From all incipient species-to-be, we pick either the youngest or oldest
species-to-be to represent the full species. This species tree we refer
to as 'the true species tree'.

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference we can use this as a prior, which we know is the truth.

The DNA sequence length and mutation rate was chosen as such to provide one
nucleotide change per 1000 years per lineage. As our time unit is in 
millions of years, this equates to a rate of 1000 nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K. As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model.

The simulated alignment is the data that could be measured in the field.
This alignment is used to infer a posterior of, using BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC chain length is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC chain length is elongated. An
MCMC chain is elongated by the proportion that is expected to increase it to 220. 
For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
would bring it to (165\%*1.33=) 220. We use
220 as a safeguard to prevent an asymptotic approach to 200.

The nLTT statistic is used to quantify the similarity of each posterior's phylogeny 
with 'the true species tree'. In this way, a distribution of phylogenies
is transformed to a distribution of nLTT statistics. With this, the generative
model of this experiment has been completely described, from parameter estimates
to data (nLTT statistics) generated.

To quantify the effect of sampling (either the youngest or oldest species-to-be to represent
a species), for each parameter set that allows for incipient species,
we simulate 1000 incipient species trees that result in different species trees
when sampled either way. From each species tree, we compare the nLTT distributions
for the two sampling methods. These distributions are compared 
using the method described in [Kruschke 2013], with a non-informative
prior. In brief: if the 95\% Highest Density Interval of the differences between 
these two distributions excludes zero, it is credible the distributions are different.

To measure if protractedness matters, we compare the nLTT distribution per
combination of speciation initiation and extinction rate for a protracted
setting with its non-protracted (with a high speciation completion 
rate) setting. So, for each speciation initiation and extinction rate, 
the nLTT statistic distribution of the 
non-protracted setting is used as the baseline. That distribution is compared
with each protracted setting. This is done the same way as with the two
different sampling methods. [NOTE: we expect nLTT to go up for increasing SCR,
this should be in the method].

[NOTE: We do have more info, from Rampal's earlier articles. Perhaps this 
info can be put in]

[NOTE: as we know the expected number of lineages through time for both BD and PBD
model, we expect it to fall with the 95\% Highest Density Interval. It would also
increase support for Rampal's choice of parameter values (next to their 
computational feasability)]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{itemize}  
  \item 12 simulation parameters:
  \begin{enumerate}  
    \item speciation initiation rate
    \item speciation completion rate
    \item extinction rate
    \item crown age
    \item crown age sigma
    \item sampling method
    \item mutation rate
    \item sequence length
    \item mcmc length
    \item tree sim rng seed
    \item alignment rng seed
    \item beast2 rnd seed
  \end{enumerate}
  \item number of taxa in incipient species tree (also: number of subspecies)
  \item would there be a difference in brach lengths between the two sampling methods?
  \item number of taxa in species tree (also: number of species)
  \item mean nLTT statistic
  \item median nLTT statistic
  \item lower bound of 95\% HPD value of nLTT statistic
  \item upper bound of 95\% HPD value of nLTT statistic
  \item 11 ESSes of all parameters estimated by BEAST2:
  \begin{enumerate}  
    \item posterior
    \item likelihood 
    \item prior 
    \item treeLikelihood 
    \item TreeHeight 
    \item BirthDeath 
    \item BDBirthRate 
    \item BDDeathRate 
    \item logP.mrca 
    \item mrcatime 
    \item clockRate
  \end{enumerate}
\end{itemize}

>>>>

This exploratory research's goal is to produce a 
usefull and balanced data set, that quantifies the inference error 
we make from choosing an overly simplistic prior. In this context, 
we measure the error made on phylogenies in which speciation takes 
time, when using a Birth-Death prior.

The parameter space the data set explores is inspired on 
[Etienne et al, 2014] for two reasons: (1) the academic
reason of being able to compare results, and (2) the practical
reason that the simulated phylogies will have enough taxa to
be informative, but not too big to be computationally overdemanding.
This research uses a complete superset of those parameter settings.
For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit, 
extending the set of
values used from one ($0.5$) to three. 
The speciation completion
rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
events per (incipient species) lineage per time unit, in which only
the last value is added as a computational proxy for infinity, which
makes the experiment fall back to a Birth-Death model (as speciation
is again instantaneous), allowing to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit,
in which $0.4$ is a new extinction rate.
We investigate these parameters in a factorial fashion, except for
(1) those combinations in which the extinction rate equals or exceeds
the speciation initiation rate, and (2) the difference between
speciation initiation rate and extinction rate exceeds 0.8. This is
done to prevent overly taxon-poor and taxon-rich phylogenies respectively.

From each biological parameter set, a protracted birth-death tree is simulated,
with the same crown age as [Etienne et al., 2014] of 15 million years. 
Each protracted birth-death tree uses a different random number
generatior seed, and thus will be unique, resulting in a balanced 
data set.

From an incipient species tree, we sample a species tree
by picking either the youngest or oldest
species-to-be to represent the full species.
We use either sampling method in 50\% of all cases. We chose not to use
a random species-to-be to represent a full species as to maximize variance
in this aspect. Note that only in few cases (when speciation completion rate
is low) the sampling method has an effect on the branch length distribution
of the sampled species tree, as this requires a very specific scenario.

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide one
nucleotide change per 1000 years per lineage. As our time unit is in 
millions of years, this equates to a rate of 1000 nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K. As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.

From an alignment, we run a Bayesian analysis and create a posterior, 
using the phylogetic tool BEAST2. We assume a
JC69 nucleotide substitution (as we do use that), a strict clock model (as
we have that) and a Birth-Death tree prior (as this simplification is the goal
of this research). Additionally, we assume a MRCA prior with a normal distribution
with a mean of the crown age, and a standard deviation of 0.001 time units. 0.001
time units equates to 1000 years, which is the same resolution as which the
alignment is created. After discarding a burn-in of 10\%, the MCMC algorithm must 
result in an ESS of the posterior of at least 200 and the MCMC chain length is 
chosen as such that this is the case for all simulations. For simulations
with an ESS below 200, their MCMC chain length is elongated. An
MCMC chain is elongated by the proportion that is expected to increase it to 220. 
For example, would an ESS be 165, increasing its MCMC chain length by 33\% 
would bring it to (165\%*1.33=) 220. We use 220 as a safeguard to prevent 
an asymptotic approach to 200.

We realize there will be an unavoidable bias in our data set: the MCMC
algorithm will produce phylogenies that are representative for the joint
likelihood of (correct) site model, (correct) clock model and 
(incorrect) (Birth-Death) speciation model. This will
warp the distribution of phylogenies somewhat towards the incorrect
speciation model: branch length distributions that are likely in the
Birth-Death model will be relatively overrepresented. We expect
this bias to be systematic, equally affecting distributions
of identical speciation completion rates, and symptomatic
of using contemporary tools.

Each posterior's phylogeny is compared to the 'true' species tree.
To quantify the difference between two phylogenies, the nLTT statistic
is used. The nLTT statistic equates to the area between the normalized
lineages-through-time-plots of two phylogenies, which will be a value between 
zero (for identical phylogenies) and one. As the nLTT statistic increases
when phylogenies get increasingly different, we use the nLTT statistic
as a measure of inference error. Comparing the one 'true' species tree
with each of the posterior's species trees, a distribution nLTT statistics
is created.

We predict that nLTT statistic values increase 
with an increasing protractednedness (that is, a low speciation completion rate),
but we cannot predict the the extent of this error, as it has never been 
measured [NOTE: this may be easy to change: Etienne & Rosindel 2012 show the equations
for the expected number of lineages through time. I think I may be able to pull
of a prediction here, by fitting a BD curve on a PBD curve with a minimized error,
but research would be needed to investigate that].

The data set produced by this exploratory research will consist out of two files:
one containing raw data, the other summarized data. 
A data file should be convenient to handle, thus as an upper limit for this data set's size, 
we use the maximum size of a table (called a data frame) of the R programming 
language. Within R, such a table may have maximally $2^{32}-1$ cells and occupy
less then 2 gigabytes in memory. 
The raw data file will contain parameters and 1000 nLTT statistics values.
Each row will contain the parameter set and these nLTT statistics.
From the constraints of the R programming language, this allows for 40K rows. 
The file with the summarized data will provide summary statistics of the raw
data. Each row will contain a parameter set, summary statistics and other
measurements. See the appendix, chapter 'Data set specifications', for the full list.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Glossary}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Please supply, as a separate list, the definitions of field-specific terms used in your article.

@@

\begin{itemize}  
  \item 12 simulation parameters:
  \begin{enumerate}  
    \item speciation initiation rate
    \item speciation completion rate
    \item extinction rate
    \item crown age
    \item crown age sigma
    \item sampling method
    \item mutation rate
    \item sequence length
    \item mcmc length
    \item tree sim rng seed
    \item alignment rng seed
    \item beast2 rnd seed
  \end{enumerate}
  \item expected speciation completion time (from Etienne & Rosindell 2012)
  \item expected duration of speciation (from Etienne & Rosindell 2012)
  \item number of taxa in incipient species tree (also: number of subspecies)
  \item would there be a difference in brach lengths between the two sampling methods?
  \item number of taxa in species tree (also: number of species)
  \item mean nLTT statistic
  \item median nLTT statistic
  \item lower bound of 95\% HPD value of nLTT statistic
  \item upper bound of 95\% HPD value of nLTT statistic
  \item 11 ESSes of all parameters estimated by BEAST2:
  \begin{enumerate}  
    \item posterior
    \item likelihood 
    \item prior 
    \item treeLikelihood 
    \item TreeHeight 
    \item BirthDeath 
    \item BDBirthRate 
    \item BDDeathRate 
    \item logP.mrca 
    \item mrcatime 
    \item clockRate
  \end{enumerate}
  \item [some] MCMC operator acceptances
\end{itemize}
